The illusory promise of the Aligned Rank Transform

### 2025-07-01 text/30-assumptions.qmd
tiny change

@fig-art-binary illustrates this problem using a simple dataset with a binary variable. In our example, all but one value are zero. Notice how dramatically ART alters the ranks and exaggerates the difference between the two levels of $B$ ($b_1$ vs. $b_2$). This issue persists even when a large number of participants are added and there is only a single value of 1 among thousands of 0s, causing the method to produce arbitrarily low *p*-values. Then, a single value variation (e.g., turning this 1 to 0) will cause ART’s results to drastically change. We will show that this unstable behavior makes ART inflate error rates even for variables with a larger range of discrete values.

>>>>

@fig-art-binary illustrates this problem using a simple dataset with a binary variable. In our example, all but one value are zero. Notice how dramatically ART alters the ranks and exaggerates the difference between the two levels of $B$ ($b_1$ vs. $b_2$). This issue persists even when a large number of participants are added and there is only a single value of 1 among thousands of 0s, causing the method to produce arbitrarily low *p*-values. Then, a single value variation (e.g., turning this 1 to 0) will cause ART’s results to drastically change. Although this example illustrates an extreme case, ART's instability can manifest in various forms, inflating error rates even when responses span a wider range of discrete values.
